Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation

### 2023-04-27 results.tex
Updated Results with tol 1e-4

To couple FEniCSx with Bempp-cl we load a volume mesh with FEniCSx. We then export the corresponding boundary mesh into Bempp-cl and assemble the boundary spaces there. Bempp-cl provides numerical trace operators that can translate from the degree of freedom (dof) representation in FEniCSx to the dof representation in Bempp-cl. The corresponding translation work is handled opaquely and the user only needs to deal with high-level interfaces of FEniCSx operators, Bempp-cl operators, and trace operators. FMM assembly fits automatically into this framework and can be enabled or disabled as a simple configuration option.

@@

In implicit-solvent models, the molecule is usually considered as a region with constant permittivity, in this case, with $\epsilon_1=2$.
In the solvent region, we used the permittivity of water ($\epsilon_2$=80) and an inverse of the Debye length of $\kappa=0.125$ \AA$^{-1}$.
As in these cases there is an analytical solution for $\phi_c$ in Eq. \eqref{eq:phic}, we compute $\phi_r$ with Eq. \eqref{eq:phi_reac} in both BEM-BEM and FEM-BEM coupling approaches. Then, for FEM-BEM, the integral over $\Gamma$ in Eq. \eqref{eq:phi_reac} corresponds to the trace of the solution vector from Eq. \eqref{eq:fembem_matrix}.

@@

As a more realistic test, we assessed the performance of the FEM-BEM coupling technique against a purely BEM implementation for arginine.
The structure of arginine taken from the protein data bank, and parameterized with the Amber\cite{ponder2003force} force field. 
We generated surface meshes containing 4.1, 6.7, 8.6, 17, and 24.5 vertices per \AA$^2$ with Nanoshaper.~\cite{decherchi2013general}
These densities correspond to a grid scale parameter in Nanoshaper equal to 1.6, 2.0, 2.4, 3.4, and 4.0, respectively, where the grid scale is the reciprocal of the average characteristic length of the triangles.
The surface meshes were inputs to our purely BEM solver and to create the FEM mesh with pyGAMer,~\cite{lee2020open} which invoked TetGen\cite{hang2015tetgen} with a quality parameter (radius-edge ratio) of 1.0.

The solvation energy computed with the two schemes is presented by Figure \ref{fig:arg_constant_energy}, which, as expected, converges to a similar answer as the mesh is refined.
Figure \ref{fig:arg2_constant_time_iter} compares the iteration count and time-to-solution. The left plot shows that using only BEM outperforms the coupled FEM-BEM approach in terms of iterations count. However, if we look at the total time that solvers take to obtain the solution, we can see the advantage of using the FEM-BEM coupling. The higher computational cost is caused by the need of using a hypersingular operator in the BEM formulation, and the fact that we are not using any acceleration method ($ie.$ FMM). Regardless, timings for FEM-BEM coupling scale much worse with size than its pure BEM counterpart, indicating that BEM-BEM coupling is better for larger problems. 
%\textcolor{red}{Chris: I thought BEM-BEM outperformed FEM-BEM in total time. Am I wrong there?} \textcolor{blue}{Michal: Not for this formulation of BEM-BEM since we have hypersingular operator. But you can see that for denser meshes we should outperform it.}

@@

In contrast to a purely BEM approach, FEM-BEM coupling gives flexibility to consider space-varying field parameters. 
A popular description of the molecule is to consider a permittivity that varies like a Gaussian around each atom,\cite{grant2001smooth} which has shown enhanced accuracy in some applications, like pKa calculations.\cite{li2013dielectric}
In this setting, we define a density function $\rho$ depending on position $r$ as
%
\begin{equation}
\rho(r) := \prod_i \left(1 - \exp{\left(\frac{\|r-\mathbf{x}_i\|}{\sigma^2 R_i^2}\right)}\right)
\end{equation}
%
where the product runs over all the atoms of the solute, $R_i$ is the van der Waals radius of atom i, and we used $\sigma$=1. Then, we can compute the permittivity as
%
\begin{equation}\label{eq:varying_eps}
\epsilon := \left(1-\rho \right) \epsilon_1 + \rho\epsilon_2
\end{equation}

@@

Table \ref{table:arg_variable} shows a comparison of solvation energy computed with APBS and our FEM-BEM coupling approach. We can see that they are both converging to equivalent values, where the finest meshes agree up to 0.5\% (0.5 kcal/mol). The coarsest meshes used in both cases are within the recommended densities for accurate solvation energy simulations with constant permittivity. For example, a finite difference mesh with $h$=0.5 \AA~ or less is recommended for binding energy calculations\cite{sorensen2015comprehensive} (the result of substracting two solvation energies), and a similar study with BEM\cite{CooperBardhanBarba2014} showed that a mesh with 2 vertices per \AA$^2$ gives acceptable results when computing binding and solvation energies. We see a jump in the solvation energy for the FEM-BEM results going from 4.1 to 6.7 vertices per \AA$^2$, but later the energy monotonically decreases, converging to a solution. This is an indication that mesh requirements for variable permitivities may be tighter than with a constant permittivity, even though the jump in dielectric constant accross the molecular surface is usually smaller.

%Figure \ref{fig:arg2_variable} contains the performance results of this test case, where we distinguish the iteration count from the calculation in dissolved and vacuum states. \textcolor{red}{Chris: if we don't include Fig. \ref{fig:arg2_variable}, we should remove this paragraph. I believe that if we show performance for fem-bem, a reviewer might be tempted to ask it for APBS too, which might be difficult.}  \textcolor{blue}{Michal: I completely agree. We were even talking about not putting them, but just for our information, we kept it for now. They weren't going to be in the final version}
\begin{table}
\centering
\begin{tabular}{c|c|c|c}
&Mesh size  & Grid points & $\Delta G_{solv}$\\
&\AA       & &  kcal/mol \\
\hline
\multirow{4}{*}{APBS}& 0.52$\times$0.52$\times$0.52 & 97$\times$97$\times$97 & -107.6186 \\ 
& 0.39$\times$0.39$\times$0.39 & 129$\times$129$\times$129 & -107.8752\\ 
&0.26$\times$0.26$\times$0.26 & 193$\times$193$\times$193& -108.3378\\ 
&0.195$\times$0.195$\times$0.195 & 257$\times$257$\times$257& -108.5837\\ 
&0.098$\times$0.098$\times$0.098 & 513$\times$513$\times$513& -108.8844\\ 
\hline
&Mesh dens. & DOFs & $\Delta G_{solv}$\\
&vert/\AA$^2$   &  &  kcal/mol \\
\hline
%\multirow{5}{*}{Standard FEM-BEM}& 2 & -36.239\\
    & 4.1 & 3 491 & -109.933 \\
FEM-BEM    & 6.7  & 5 787 & -110.238 \\
coupling    & 8.57  & 8 844 & -109.658 \\
    & 17 & 19 911 & -109.368 \\
    & 24.5 & 32 302 & -109.286 \\
\hline
%\multirow{5}{*}{Standard FEM-BEM}& 2 & -29.554\\
%    & 2 & -34.670\\
%Hybrid    & 4  & -33.603 \\
%FEM-BEM    & 8  & -32.822 \\
%    & 16 & -32.072 \\
%\hline
\end{tabular}
\caption{Solvation energy of arginine with a Gaussian-like permittivity, computed using the FEM-BEM approach and APBS. The mesh density for FEM-BEM corresponds to the vertex density of the surface mesh used to generate the volumetric mesh.}
\label{table:arg_variable}
\end{table}

@@

So far, we have only tested the FEM-BEM coupled approach with small structures. Here, we study its behaviour with larger structures to evaluate its applicability in more realistic settings, and  in test cases where BEM-BEM coupling is not an alternative. Using the same Gaussian permittivity field inside the molecule, we computed the solvation free energy of protein G B1 (955 atoms, PDB code 1pgb), lysozyme (1960 atoms, PDB code 1lyz), the barnase-barstar complex (9464 atoms, PDB code 1x1u), and immunoglobulin G (20148 atoms, PDB code 1igt). All structures were parameterized with the Amber\cite{Swanson05} force field using pdb2pqr.\cite{Dolinsky04} The surface was meshed with Nanoshaper\cite{decherchi2013general} considering a grid scale of 1.5, and then used to mesh the solute region with pyGAMer\cite{lee2020open} and TetGen\cite{hang2015tetgen} with a radius-edge ratio of 1.0. Solvation free energy and timings for these runs are presented in Table \ref{table:large_variable}, showing that our FEM-BEM coupling method can reach medium-to-large sized proteins on a workstation. For these larger structures we enabled the FMM capabilities of Bempp-cl.

Table \ref{table:large_variable} shows that the iteration count increases with the problem size. This is expected for Johnson-N\'ed\'elec, which is the simplest coupling strategy. To isolate the effect of the increased iterations in the analysis, we separate timings in ``setup'' and ``solving'' time, where the ``setup'' time is independent of the number of iterations, and the ``solving'' time corresponds to the time spent in the GMRES solver. Then, the time-per-iteration in Table \ref{table:large_variable} only considers the ``solving'' time. Looking at the increase in degrees of freedom (DOFs), which grows equivalently for FEM and BEM, the ``setup'' time scales slightly worse than $\mathcal{O}(N^2)$, but represents the smalle fraction of the total time. On the other hand, the ``solving time per iteration'' scales closer to $\mathcal{O}(N)$, which is expected, as we are using a FMM for the BEM portion of the matrix. This is an indication that the high ``solving time'' is mainly due to the increase in iteration count, and having better conditioned coupling methods, such as the so-called hybrid approach,\cite{betcke2022hybrid} can have a large impact in the time to solution.

Even though this scheme is capable of calculating the electrostatic potential in medium-to-large proteins, the largest test case in Table \ref{table:large_variable} (1igt) used up 30\% of the available RAM memory. 
%(\textcolor{red}{check and rewrite \textcolor{blue}{we are using around 50\% of memory}}). 
If we were aiming at larger structures, such as full viruses,\cite{MartinezETal2019,wang2021high} we would require to improve not only in the conditioning of the system, but also in the optimization of our fast algorithms\cite{wang2021exafmm,kailasa2023pyexafmm} and parallelization.

\begin{table}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{c|c|c|c|c|c|c|c|c}
Molecule & FEM  & BEM &  $\Delta G_{solv}$ & Iterations & Setup Time & Solving Time & Solv. time per & Total Time \\
& DOFs & DOFs &  kcal/mol & & [sec] & [sec] & iter. [sec] & [sec] \\
\hline
 1pgb  & 29 434 &  10 058 &  -300.888 & 826 & 82 & 547 & 0.6 & 6 629  \\
 1lyz  & 56 114 & 18 606 &  -599.642 & 1 285 & 330 & 1 640  & 1.2 & 1 970  \\
 %1lyz gs3 & 257 391 & 75 048 &  -600.637 & 2 949  & 1 690 & 10 900 & 3.6 & 12 590   \\
 1x1u & 263 181 & 81 258 &  -1 984.468 & 4 167 & 8 240 & 16 300 & 3.91 & 24 540  \\
 1igt & 597 575 & 187 712 &   -3 294.230 & 8 361 & 41 200 & 64 200 & 7.67  &  105 400   \\
 \hline
\end{tabular}}
\caption{Results for larger structures with a variable permitivitty.}
\label{table:large_variable}
\end{table}
%
%\begin{table}
%\centering
%\begin{tabular}{c|c|c|c|c|c|c|c|c}
%Molecule & Mesh size  & FEM  & BEM &  $\Delta G_{solv}$ & Iterations & Setup Time & Solving Time & Total Time \\
%& vert/\AA$^2$ & DOFs & DOFs &  kcal/mol & & sec & sec & sec \\
%\hline
% 1pgb &  & 29 434 &  10 058 &  -300.888 & 826 & 82 & 547 & 629  \\
% 1lyz &  & 56 114 & 18 606 &  -599.642 & 1 285 & 330 & 1 640  & 1 970  \\
% 1lyz gs3 &  & 257 391 & 75 048 &  -600.637 & 2 949  & 1 690 & 10 900 & 12 590   \\
% 1uxu &  & 263 181 & 81 258 &  -1 984.468 & 4 167 & 8 240 & 16 300 & 24 540  \\
% 1igt &  & 597 575 & 187 712 &   -3 294.230 & 8 361 & 41 200 & 64 200 &  105 400   \\
% \hline
%\end{tabular}
%\caption{Large tests' results}
%\label{table:large_variable}
%\end{table}

>>>>

To couple FEniCSx with Bempp-cl we load a volume mesh with FEniCSx. We then export the corresponding boundary mesh into Bempp-cl and assemble the boundary spaces there. Bempp-cl provides numerical trace operators that can translate from the degree of freedom (DOF) representation in FEniCSx to the dof representation in Bempp-cl. The corresponding translation work is handled opaquely and the user only needs to deal with high-level interfaces of FEniCSx operators, Bempp-cl operators, and trace operators. FMM assembly fits automatically into this framework and can be enabled or disabled as a simple configuration option.

@@

In implicit-solvent models, the molecule is usually considered as a region with constant permittivity, in this case, with $\epsilon_1=2$.
In the solvent region, we used the permittivity of water ($\epsilon_2$=80) and an inverse of the Debye length of $\kappa=0.125$ \AA$^{-1}$.
As in these cases, there is an analytical solution for $\phi_c$ in Eq. \eqref{eq:phic}, we compute $\phi_r$ with Eq. \eqref{eq:phi_reac} in both BEM-BEM and FEM-BEM coupling approaches. Then, for FEM-BEM, the integral over $\Gamma$ in Eq. \eqref{eq:phi_reac} corresponds to the trace of the solution vector from Eq. \eqref{eq:fembem_matrix}.

@@

As a more realistic test, we assessed the performance of the FEM-BEM coupling technique against a purely BEM implementation for arginine.
The structure of arginine was taken from the protein data bank, and parameterized with the Amber\cite{ponder2003force} force field. 
We generated surface meshes containing 4.1, 6.7, 8.6, 17, and 24.5 vertices per \AA$^2$ with Nanoshaper.~\cite{decherchi2013general}
These densities correspond to a grid-scale parameter in Nanoshaper equal to 1.6, 2.0, 2.4, 3.4, and 4.0, respectively, where the grid scale is the reciprocal of the average characteristic length of the triangles.
The surface meshes were inputs to our purely BEM solver and to create the FEM mesh with pyGAMer,~\cite{lee2020open} which invoked TetGen\cite{hang2015tetgen} with a quality parameter (radius-edge ratio) of 1.0.

The solvation energy computed with the two schemes is presented in Figure \ref{fig:arg_constant_energy}, which, as expected, converges to a similar answer as the mesh is refined.
Figure \ref{fig:arg2_constant_time_iter} compares the iteration count and time-to-solution. The left plot shows that using only BEM outperforms the coupled FEM-BEM approach in terms of the iteration count. However, if we look at the total time that solvers take to obtain the solution, we can see the advantage of using the FEM-BEM coupling. The higher computational cost is caused by the need of using a hypersingular operator in the BEM formulation, and the fact that we are not using any acceleration method ($ie.$ FMM). Regardless, timings for the FEM-BEM coupling scale are much worse with size than its pure BEM counterpart, indicating that BEM-BEM coupling is better for larger problems. 
%\textcolor{red}{Chris: I thought BEM-BEM outperformed FEM-BEM in total time. Am I wrong there?} \textcolor{blue}{Michal: Not for this formulation of BEM-BEM since we have hypersingular operator. But you can see that for denser meshes we should outperform it.}

@@

In contrast to a purely BEM approach, FEM-BEM coupling gives the flexibility to consider space-varying field parameters. 
A popular description of the molecule is to consider a permittivity that varies like a Gaussian around each atom,\cite{grant2001smooth} which has shown enhanced accuracy in some applications, like pKa calculations.\cite{li2013dielectric}
In this setting, we define a density function $\rho$ depending on position $r$ as
%
\begin{equation}
\rho(r) := \prod_i \left(1 - \exp{\left(\frac{\|r-\mathbf{x}_i\|}{\sigma^2 R_i^2}\right)}\right)
\end{equation}
%
where the product runs over all the atoms of the solute, $R_i$ is the van der Waals radius of atom i, and we used $\sigma$=1. Then, we can compute the permittivity as
%
\begin{equation}\label{eq:varying_eps}
\epsilon := \left(1-\rho \right) \epsilon_1 + \rho\epsilon_2
\end{equation}

@@

Table \ref{table:arg_variable} shows a comparison of solvation energy computed with APBS and our FEM-BEM coupling approach. We can see that they are both converging to equivalent values, where the finest meshes agree up to 0.5\% (0.5 kcal/mol). The coarsest meshes used in both cases are within the recommended densities for accurate solvation energy simulations with constant permittivity. For example, a finite difference mesh with $h$=0.5 \AA~ or less is recommended for binding energy calculations\cite{sorensen2015comprehensive} (the result of subtracting two solvation energies), and a similar study with BEM\cite{CooperBardhanBarba2014} showed that a mesh with 2 vertices per \AA$^2$ gives acceptable results when computing binding and solvation energies. We see a jump in the solvation energy for the FEM-BEM results going from 4.1 to 6.7 vertices per \AA$^2$, but later the energy monotonically decreases, converging to a solution. This is an indication that mesh requirements for variable permittivities may be tighter than with a constant permittivity, even though the jump in dielectric constant across the molecular surface is usually smaller.

%Figure \ref{fig:arg2_variable} contains the performance results of this test case, where we distinguish the iteration count from the calculation in dissolved and vacuum states. \textcolor{red}{Chris: if we don't include Fig. \ref{fig:arg2_variable}, we should remove this paragraph. I believe that if we show performance for fem-bem, a reviewer might be tempted to ask it for APBS too, which might be difficult.}  \textcolor{blue}{Michal: I completely agree. We were even talking about not putting them, but just for our information, we kept it for now. They weren't going to be in the final version}
\begin{table}
\centering
\begin{tabular}{c|c|c|c}
&Mesh size  & Grid points & $\Delta G_{solv}$\\
&\AA       & &  kcal/mol \\
\hline
\multirow{4}{*}{APBS}& 0.52$\times$0.52$\times$0.52 & 97$\times$97$\times$97 & -107.6186 \\ 
& 0.39$\times$0.39$\times$0.39 & 129$\times$129$\times$129 & -107.8752\\ 
&0.26$\times$0.26$\times$0.26 & 193$\times$193$\times$193& -108.3378\\ 
&0.195$\times$0.195$\times$0.195 & 257$\times$257$\times$257& -108.5837\\ 
&0.098$\times$0.098$\times$0.098 & 513$\times$513$\times$513& -108.8844\\ 
\hline
&Mesh dens. & DOFs & $\Delta G_{solv}$\\
&vert/\AA$^2$   &  &  kcal/mol \\
\hline
%\multirow{5}{*}{Standard FEM-BEM}& 2 & -36.239\\
    & 4.1 & 3 491 & -109.907 \\
FEM-BEM    & 6.7  & 5 787 & -110.199 \\
coupling    & 8.57  & 8 844 & -109.582 \\
    & 17 & 19 911 & -109.350 \\
    & 24.5 & 32 302 & -109.265 \\
\hline
%\multirow{5}{*}{Standard FEM-BEM}& 2 & -29.554\\
%    & 2 & -34.670\\
%Hybrid    & 4  & -33.603 \\
%FEM-BEM    & 8  & -32.822 \\
%    & 16 & -32.072 \\
%\hline
\end{tabular}
\caption{Solvation energy of arginine with a Gaussian-like permittivity, computed using the FEM-BEM approach and APBS. The mesh density for FEM-BEM corresponds to the vertex density of the surface mesh used to generate the volumetric mesh.}
\label{table:arg_variable}
\end{table}

@@

So far, we have only tested the FEM-BEM coupled approach with small structures. Here, we study its behaviour with larger structures to evaluate its applicability in more realistic settings, and in test cases where BEM-BEM coupling is not an alternative. Using the same Gaussian permittivity field inside the molecule, we computed the solvation free energy of protein G B1 (955 atoms, PDB code 1pgb), lysozyme (1960 atoms, PDB code 1lyz), the barnase-barstar complex (9464 atoms, PDB code 1x1u), and immunoglobulin G (20148 atoms, PDB code 1igt). All structures were parameterized with the Amber\cite{Swanson05} force field using pdb2pqr.\cite{Dolinsky04} The surface was meshed with Nanoshaper\cite{decherchi2013general} considering a grid scale of 1.5, and then used to mesh the solute region with pyGAMer\cite{lee2020open} and TetGen\cite{hang2015tetgen} with a radius-edge ratio of 1.0. Solvation free energy and timings for these runs are presented in Table \ref{table:large_variable}, showing that our FEM-BEM coupling method can reach medium-to-large-sized proteins on a workstation. For these larger structures, we enabled the FMM capabilities of Bempp-cl.

Table \ref{table:large_variable} shows that the iteration count increases with the problem size. This is expected for Johnson-N\'ed\'elec, which is the simplest coupling strategy. To isolate the effect of the increased iterations in the analysis, we separate timings in ``setup'' and ``solving'' time, where the ``setup'' time is independent of the number of iterations, and the ``solving'' time corresponds to the time spent in the GMRES solver. Then, the time-per-iteration in Table \ref{table:large_variable} only considers the ``solving'' time. Looking at the increase in degrees of freedom (DOFs), which grows equivalently for FEM and BEM, the ``setup'' time scales slightly worse than $\mathcal{O}(N^2)$, but represents the smaller fraction of the total time. On the other hand, the ``solving time per iteration'' scales closer to $\mathcal{O}(N)$, which is expected, as we are using an FMM for the BEM portion of the matrix. This is an indication that the high ``solving time'' is mainly due to the increase in iteration count, and having better-conditioned coupling methods, such as the so-called hybrid approach,\cite{betcke2022hybrid} can have a large impact on the time to solution.

Even though this scheme is capable of calculating the electrostatic potential in medium-to-large proteins, the largest test case in Table \ref{table:large_variable} (1igt) used up 30\% of the available RAM memory. 
%(\textcolor{red}{check and rewrite \textcolor{blue}{we are using around 50\% of memory}}). 
If we were aiming at larger structures, such as full viruses,\cite{MartinezETal2019,wang2021high} we would require to improve not only in the conditioning of the system but also in the optimization of our fast algorithms\cite{wang2021exafmm,kailasa2023pyexafmm} and parallelization.

\begin{table}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{c|c|c|c|c|c|c|c|c}
Molecule & FEM  & BEM &  $\Delta G_{solv}$ & Iterations & Setup Time & Solving Time & Solv. time per & Total Time \\
& DOFs & DOFs &  kcal/mol & & [sec] & [sec] & iter. [sec] & [sec] \\
\hline
 1pgb  & 29 434 &  10 058 &  -300.389 & 613 & 89 & 410 & 0.67 & 702  \\
 1lyz  & 56 114 & 18 606 &  -597.515 & 724 & 334 & 921  & 1.27 & 1 255   \\
 %1lyz gs3 & 257 391 & 75 048 &  -600.637 & 2 949  & 1 690 & 10 900 & 3.6 & 12 590   \\
 1x1u & 263 181 & 81 258 &  -1 975.068 & 1 911 & 8 240 & 7 410 & 3.88 & 15 650  \\
 1igt & 597 575 & 187 712 &   -3 279.907 & 1 974 & 40 930 & 15 600 & 7.9  &  56 530   \\
 \hline
\end{tabular}}
\caption{Results for larger structures with a variable permittivity.}
\label{table:large_variable}
\end{table}
%
%\begin{table}
%\centering
%\begin{tabular}{c|c|c|c|c|c|c|c|c}
%Molecule & Mesh size  & FEM  & BEM &  $\Delta G_{solv}$ & Iterations & Setup Time & Solving Time & Total Time \\
%& vert/\AA$^2$ & DOFs & DOFs &  kcal/mol & & sec & sec & sec \\
%\hline
% 1pgb &  & 29 434 &  10 058 &  -300.888 & 826 & 82 & 547 & 629  \\
% 1lyz &  & 56 114 & 18 606 &  -599.642 & 1 285 & 330 & 1 640  & 1 970  \\
% 1lyz gs3 &  & 257 391 & 75 048 &  -600.637 & 2 949  & 1 690 & 10 900 & 12 590   \\
% 1uxu &  & 263 181 & 81 258 &  -1 984.468 & 4 167 & 8 240 & 16 300 & 24 540  \\
% 1igt &  & 597 575 & 187 712 &   -3 294.230 & 8 361 & 41 200 & 64 200 &  105 400   \\
% \hline
%\end{tabular}
%\caption{Large tests' results}
%\label{table:large_variable}
%\end{table}
